---
title: "Complete Workflow for Field Key Generation"
author: "INBO BWK Field Keys Project"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Workflow for Field Key Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE to prevent actual execution during build
)
```

## Introduction

This vignette demonstrates the complete workflow for generating interactive HTML field identification keys for habitat classification (BWK - Biologische Waarderingskaart). The workflow reads structured data from Google Sheets, processes it, validates the logic, and generates both HTML and PDF outputs.

## Overview

The field key system supports identification of 5 main habitat types:

1. **Bos** (bos) - Forest key
2. **Grasland** (grl) - Grassland key  
3. **Heide** (hei) - Heathland key
4. **Moeras** (mrs) - Wetland key
5. **Water** (wtr) - Water key

Each key consists of numbered steps with questions that guide users to habitat classifications according to BWK and Natura 2000 standards.

## Prerequisites

### Required Packages

```{r packages}
library(tidyverse)      # Data manipulation
library(googlesheets4)  # Reading Google Sheets
library(glue)           # String interpolation
library(yaml)           # YAML export
```

### Authentication

Before reading data from Google Sheets, you need to authenticate:

```{r auth}
# Replace with your email address
gs4_auth(email = "your.email@inbo.be")
```

### Source Functions

All functions are located in the `source/functions` directory:

```{r source-functions}
# Source all function files
fun_path <- file.path("source", "functions")
all_function_files <- list.files(
  path = fun_path,
  pattern = "^[^_].*\\.[rR]$",
  full.names = TRUE
)

status <- lapply(all_function_files, FUN = source)
```

## Workflow Steps

### Step 1: Define Field Keys

Define the Google Sheets containing your field key data:

```{r define-keys}
sleutel_df <- tribble(
  ~habitat, ~sheet_id, ~afkorting,
  "bos", "11_sh5YYlc5RFY8v7ECecAlm-4Euo_4ph2CfIPWb6CBw", "bos",
  "grasland", "1dgSVnJBSMm4G4nPNriDzfB1XhyV0xfo-1yqcgjWwr20", "grl",
  "heide", "1d38qnUQXfEvkee0HtlI_L8aC76myZdSSAiWg_fLIEh0", "hei",
  "moeras", "1eKtmWcgURTCZvshOQPS_IDOga8fC3ADmJ8-p4_rTkHo", "mrs",
  "water", "17o7UHZipArEBwQiHKxIqHmnaL2YB84Y4x1dHkgSMDBk", "wtr"
)
```

Each row specifies:

- `habitat`: Full habitat name
- `sheet_id`: Google Sheets document ID
- `afkorting`: Abbreviated code used internally

### Step 2: Read Data from Google Sheets

The `read_gsheet_data()` function reads all sheets and returns a named list:

```{r read-data}
# Read raw data from Google Sheets
veldsleutels_raw <- read_gsheet_data(sleutel_df)

# Optionally save for offline work
saveRDS(veldsleutels_raw, file.path("interim", "veldsleuteldata.RDS"))

# Load saved data
veldsleutels_raw <- readRDS(file.path("interim", "veldsleuteldata.RDS"))
```

**What happens**: For each habitat, the function:

- Reads metadata from cells A1:B4
- Reads key data from cells A5:H1000
- Adds the abbreviated key code to each row
- Returns a list with one data frame per habitat

### Step 3: Prepare Data

The `prepare_data()` function validates and transforms the raw data:

```{r prepare-data}
# Prepare all keys
veldsleutels_prep <- lapply(veldsleutels_raw, prepare_data)

# Check dimensions
sapply(veldsleutels_prep, dim)
```

**What happens**: This function:

- Validates required columns are present
- Filters out invalid rows
- Maps type codes (T1 → h2, Q → question, A → answer, etc.)
- Creates internal step IDs (e.g., "bos0001")
- Calculates incoming steps (which steps link to each step)
- Replaces NA values with empty strings

**Data structure after preparation**:

```{r data-structure-prep, echo=FALSE}
# Example of prepared data structure:
# STEP  TYPE  NAME                NEXT_STEP  .type      .step     .nextstep
# 1     T1    "Bostype bepalen"   NA         "h2"       "bos0001" ""
# 1     Q     "Zijn er bomen?"    NA         "question" "bos0001" ""
# 1     A     "Ja"                2          "answer"   "bos0001" "bos0002"
# 1     A     "Nee"               5          "answer"   "bos0001" "bos0005"
# 2     T1    "Loofbos?"          NA         "h2"       "bos0002" ""
# 2     Q     "Overwegend loof?"  NA         "question" "bos0002" ""
```

### Step 4: Parse Data

The `parse_data()` function converts tabular data to hierarchical structure:

```{r parse-data}
# Parse into nested lists
veldsleutels <- lapply(veldsleutels_prep, parse_data)

# Optionally export to YAML for inspection
veldsleutels_yaml <- yaml::as.yaml(veldsleutels)
cat(veldsleutels_yaml, file = "veldsleutels.yaml")
```

**What happens**: Creates a hierarchical structure where:

- Each step becomes a list element keyed by internal ID
- Headers, questions, and background info are nested
- Answers are collected in arrays with navigation links
- Info rows are attached to their preceding answer

**Example parsed structure**:

```{r parsed-structure, echo=FALSE}
# veldsleutels[["bos"]][["bos0001"]] contains:
# $step_number: "1"
# $html_id: "bos0001"
# $incoming_step: list()
# $h2: list(name = "Bostype bepalen", remark = "")
# $question: list(name = "Zijn er bomen?", remark = "")
# $answer: list(
#   list(name = "Ja", nextstep = "bos0002", n2000 = "", bwk = "", ...),
#   list(name = "Nee", nextstep = "bos0005", n2000 = "", bwk = "", ...)
# )
```

### Step 5: Validate Logic

The `check_logic()` function validates the key structure:

```{r check-logic}
check_logic(veldsleutels)
```

**What it checks**:

1. **Stale Answers**: Identifies answers that:
   - Have no next step AND
   - Have no result codes (n2000, bwk, or otherkey)
   - These represent dead ends that should be fixed

2. **Unreachable Steps**: Identifies steps that:
   - Are never referenced as a 'nextstep' by any answer
   - Cannot be reached by users
   - Except the first step (entry point)

**Example output**:

```
--- Analysing Key Group: BOS ---
[OK] All steps are reachable.

--- Analysing Key Group: GRASLAND ---
[STALE ANSWER] Step 5, Answer 2 ('Onzeker') is a dead end.
[UNREACHABLE STEPS]: The following step numbers are never called by an answer:
12, 15
```

### Step 6: Generate HTML Pages

Generate interactive HTML pages for each key:

```{r generate-html}
# Get list of all keys
main_keys <- names(veldsleutels)  # c("bos", "grasland", "heide", "moeras", "water")

# Generate individual key pages
for (k in main_keys) {
  html_write_key_page(k, veldsleutels[[k]], folder = "www")
}

# Generate index page
html_write_index_page(main_keys, folder = "", subfolder = "www")
```

**What is created**:

- **index.html**: Main landing page with buttons for each key
- **www/bos.html**: Forest identification key
- **www/grasland.html**: Grassland identification key
- **www/heide.html**: Heathland identification key
- **www/moeras.html**: Wetland identification key
- **www/water.html**: Water identification key

**HTML features**:

- Responsive design with card-based layout
- Clickable answer buttons that navigate between steps
- Color-coded result boxes for terminal answers
- Back links showing which steps can reach the current step
- Additional information panels (info, remarks, related keys)

### Step 7: Generate PDF Documentation

There are two options for generating PDF documentation:

#### Option A: Traditional R Markdown (Combined Document)

Create a single PDF with all keys:

```{r generate-pdf-rmd}
# Create R Markdown file
create_pdf_markdown(veldsleutels, rmdfile = "veldsleutels.Rmd")

# Render to PDF
rmarkdown::render("veldsleutels.Rmd")
```

**What is created**:

- **veldsleutels.Rmd**: R Markdown source file
- **veldsleutels.pdf**: Compiled PDF with all keys

#### Option B: Quarto with Flandersqmd Style (Individual Documents)

Create separate PDFs for each key with INBO corporate identity:

```{r generate-pdf-quarto}
# Create Quarto documents
create_quarto_pdf(veldsleutels, output_dir = "pdf_output", use_flandersqmd = TRUE)

# Render individual keys to PDF using Quarto
# (requires Quarto to be installed: https://quarto.org)
main_keys <- names(veldsleutels)
for (key in main_keys) {
  system(sprintf("quarto render pdf_output/%s.qmd", key))
}

# Or render all at once
system("cd pdf_output && quarto render")
```

**What is created**:

- **pdf_output/bos.qmd**: Quarto source for forest key
- **pdf_output/grl.qmd**: Quarto source for grassland key  
- **pdf_output/hei.qmd**: Quarto source for heathland key
- **pdf_output/mrs.qmd**: Quarto source for wetland key
- **pdf_output/wtr.qmd**: Quarto source for water key
- **pdf_output/README.md**: Instructions for rendering
- **PDF files**: One per key with INBO styling

**Flandersqmd features**:

- INBO/Flanders corporate identity styling
- Professional cover pages
- Formatted callout boxes for background info and results
- Cross-references with hyperlinks
- Table of contents
- Proper typography and spacing

To use flandersqmd styling, install the Quarto extension first:
```bash
quarto install extension inbo/flandersqmd-book
```

**PDF features**:

- Table of contents with clickable links
- One section per habitat key
- Internal cross-references between steps
- Formatted questions, answers, and results
- Page breaks between major sections

## Data Structure Requirements

### Google Sheets Format

Each Google Sheet should have:

**Metadata (cells A1:B4)**:

```
Field       Value
Title       Bossleutel
Version     2.0
Author      INBO
Date        2024-01-15
```

**Data (starting at row 5)**:

| STEP | TYPE | NAME | NEXT_STEP | CLASSIFICATION | BWK_CODE | SUBKEY | REMARK | KEY |
|------|------|------|-----------|----------------|----------|--------|--------|-----|
| 1 | T1 | Bostype | | | | | | bos |
| 1 | Q | Zijn er bomen aanwezig? | | | | | | bos |
| 1 | A | Ja | 2 | | | | | bos |
| 1 | A | Nee | 5 | | | | | bos |
| 2 | T1 | Loofbos of naaldbos? | | | | | | bos |
| 2 | Q | Overwegend loofhout? | | | | | | bos |
| 2 | A | Ja | 3 | | | | | bos |
| 2 | A | Nee | 4 | | | | | bos |
| 3 | T1 | Loofbos classificatie | | | | | | bos |
| 3 | Q | Welk type loofbos? | | | | | | bos |
| 3 | A | Eikenbos | | 9120 | lhb | | | bos |
| 3 | A | Beukenbos | | 9130 | lhc | | | bos |

### Column Descriptions

- **STEP**: Step number (must be numeric)
- **TYPE**: Row type code
  - `T1`, `T2`, `T3`: Headers (levels 1-3)
  - `Q`: Question
  - `A`: Answer
  - `I`: Additional info (attached to previous answer)
  - `BI`: Background information
- **NAME**: Display text for the row
- **NEXT_STEP**: Step number to navigate to (for answers)
- **CLASSIFICATION**: Natura 2000 habitat code (for terminal answers)
- **BWK_CODE**: BWK classification code (for terminal answers)
- **SUBKEY**: Reference to another key (for answers that redirect)
- **REMARK**: Additional notes or clarifications
- **KEY**: Abbreviated key code (e.g., "bos", "grl")

## Output Examples

### HTML Interface

The HTML output provides an interactive web interface:

1. **Index Page**: Grid of buttons for each habitat type
2. **Key Pages**: Step-by-step navigation with:
   - Question prompts
   - Clickable answer buttons
   - Automatic navigation between steps
   - Result displays for classifications
   - Background information panels

### PDF Document

The PDF output provides a printable reference:

**Option A (R Markdown):**
- Combined document with all keys
- Table of contents for easy navigation
- Cross-references between steps
- Basic PDF formatting

**Option B (Quarto with flandersqmd):**
- Individual PDFs for each key
- INBO/Flanders corporate identity styling
- Professional cover pages and formatting
- Enhanced callout boxes and visual elements
- Better typography and spacing

## Customization

### Styling

To customize the HTML appearance, modify `html_get_style()`:

```{r custom-style, eval=FALSE}
# Edit source/functions/html_get_style.R
# Adjust colors, fonts, sizes, etc.
```

### Output Locations

Control where files are generated:

```{r custom-output, eval=FALSE}
# HTML in custom directory
html_write_key_page("bos", veldsleutels[["bos"]], folder = "output/html")

# Index in custom location
html_write_index_page(main_keys, folder = "public", subfolder = "keys")

# PDF with custom name or location
create_pdf_markdown(veldsleutels, rmdfile = "my_keys.Rmd")

# Quarto PDFs with custom settings
create_quarto_pdf(veldsleutels, 
                  output_dir = "custom_output",
                  author_name = "Custom Author",
                  use_flandersqmd = FALSE)
```

## Troubleshooting

### Common Issues

**Authentication errors**:
```{r troubleshoot-auth, eval=FALSE}
# Clear authentication and re-authorize
gs4_deauth()
gs4_auth(email = "your.email@inbo.be")
```

**Missing columns error**:
```
Error: all columns must be present
```
- Check that your Google Sheet has all required columns
- Column names are case-sensitive

**Undefined TYPE error**:
```
Error: undefined TYPE found
```
- Check for typos in the TYPE column
- Valid types: T1, T2, T3, Q, A, I, BI

**Stale answers or unreachable steps**:
- Review output from `check_logic()`
- Verify NEXT_STEP values are correct
- Ensure terminal answers have classification codes

## Complete Example Script

Here's a minimal complete script:

```{r complete-example}
# 1. Setup
library(tidyverse)
library(googlesheets4)
library(glue)

# 2. Source functions
fun_path <- file.path("source", "functions")
all_function_files <- list.files(
  path = fun_path,
  pattern = "^[^_].*\\.[rR]$",
  full.names = TRUE
)
lapply(all_function_files, source)

# 3. Authenticate
gs4_auth(email = "your.email@inbo.be")

# 4. Define keys
sleutel_df <- tribble(
  ~habitat, ~sheet_id, ~afkorting,
  "bos", "11_sh5YYlc5RFY8v7ECecAlm-4Euo_4ph2CfIPWb6CBw", "bos"
)

# 5. Process data
veldsleutels_raw <- read_gsheet_data(sleutel_df)
veldsleutels_prep <- lapply(veldsleutels_raw, prepare_data)
veldsleutels <- lapply(veldsleutels_prep, parse_data)

# 6. Validate
check_logic(veldsleutels)

# 7. Generate outputs
main_keys <- names(veldsleutels)

# HTML output
for (k in main_keys) {
  html_write_key_page(k, veldsleutels[[k]], folder = "www")
}
html_write_index_page(main_keys, folder = "", subfolder = "www")

# PDF output - choose one approach:

# Option A: Traditional R Markdown (all keys in one PDF)
create_pdf_markdown(veldsleutels, rmdfile = "veldsleutels.Rmd")
rmarkdown::render("veldsleutels.Rmd")

# Option B: Quarto with flandersqmd style (individual PDFs per key)
create_quarto_pdf(veldsleutels, output_dir = "pdf_output")
# Then render with: system("cd pdf_output && quarto render")
```

## Conclusion

This workflow provides a complete system for generating interactive field identification keys from structured Google Sheets data. The process is:

1. **Flexible**: Easy to update keys by editing Google Sheets
2. **Validated**: Automatic logic checking catches errors
3. **Multi-format**: Generates both interactive HTML and printable PDF (R Markdown or Quarto)
4. **Professional**: Optional INBO corporate identity styling for PDFs via flandersqmd
5. **Maintainable**: Well-documented functions for each step

For questions or issues, consult the function documentation using `?function_name` or review the source code in `source/functions/`.
